A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-30 paper/writeup.tex
typos

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulations}

Our new method \textit{extend haplotypes} is applicable to any ARG.
However its accuracy depends on the overall structure of the ARG it is applied too. 
Additionally, as \textit{ARG Robinson-Foulds}
is given as a proportion of node spans
it's value is also dependent on the pair of input ARGs. 
Thus to understand how well our methods can infer ancestral haplotypes we 
work with simulated tree sequences with a large variety in parameter choice.
To do this, we simulate ARGs containing full haplotypes using \msprime{}
\citep{kelleher2016efficient,baumdicker2021efficient},
with the \texttt{coalescing\_segments\_only} option set to False.
Although \msprime{} simulates many events that do not create a coalescence in some local tree,
it only outputs information for nodes which contain a coalescence
(i.e., are the MRCA of some pair of samples at some point on the genome).
Furthermore, by default it only outputs those segments of the genome
on which there is a coalescence.
Said another way, by default all ancestral nodes in the ARG
output by \msprime{} are the MRCA of some pair of samples at every point in the genome
on which they are represented.
However, here we are interested in those segments of genome
on which the nodes are \emph{not} coalescent;
i.e., where they are unary in the local trees.
Setting \texttt{coalescing\_segments\_only} to False includes just this information:
any ancestral segments for which these coalescent nodes
are ancestral to any samples -- so, the unary portions of their spans as well.
However, this includes more information than we want:
we hope to recover those portions of ancestral haplotypes
on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
coalesces with another spanning $[b, d)$, with $a < b < c < d$,
then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
to extend the node's span to $[a, b)$ and $[c, d)$
(on which the node is unary).
However, following this example, the first lineage might carry a segment $[x, y)$
that is disjoint from the segment $[a, d)$.
We call these segments ``isolated non-coalescent segments'';
they have also been called ``trapped unary spans'' \citep[by][]{wong2024general}.
Such isolated segments will not be recovered by our algorithm,
and would likely be unrecoverable by any other method.
So, after simulation, we first remove these isolated, non-coalescent segments.
To give an idea of what proportion of the full spans of ancestral nodes
these isolated non-coalescent segments represent,
a simulation of 1000 samples
with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
has about half the total span of all nodes in isolated, non-coalescent segments.
For more discussion of these segments, see \citet{baumdicker2021efficient}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%--------------------------------
\subsection{Simulations}

Our new method \textit{extend haplotypes} is applicable to any ARG.
However its accuracy depends on the overall structure of the ARG it is applied to. 
Additionally, as \textit{ARG Robinson-Foulds}
is given as a proportion of node spans
its' value is also dependent on the pair of input ARGs. 
Thus to understand how well our methods can infer ancestral haplotypes we 
work with simulated tree sequences with a large variety in parameter choice.
To do this, we simulate ARGs containing full haplotypes using \msprime{}
\citep{kelleher2016efficient,baumdicker2021efficient},
with the \texttt{coalescing\_segments\_only} option set to False.
Although \msprime{} simulates many events that do not create a coalescence in some local tree,
it only outputs information for nodes which contain a coalescence
(i.e., are the MRCA of some pair of samples at some point on the genome).
Furthermore, by default it only outputs those segments of the genome
on which there is a coalescence.
Said another way, by default all ancestral nodes in the ARG
output by \msprime{} are the MRCA of some pair of samples at every point in the genome
on which they are represented.
However, here we are interested in those segments of genome
on which the nodes are \emph{not} coalescent;
i.e., where they are unary in the local trees.
Setting \texttt{coalescing\_segments\_only} to False includes just this information:
any ancestral segments for which these coalescent nodes
are ancestral to any samples -- so, the unary portions of their spans as well.
However, this includes more information than we want:
we hope to recover those portions of ancestral haplotypes
on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
coalesces with another spanning $[b, d)$, with $a < b < c < d$,
then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
to extend the node's span to $[a, b)$ and $[c, d)$
(on which the node is unary).
However, following this example, the first lineage might carry a segment $[x, y)$
that is disjoint from the segment $[a, d)$.
We call these segments ``isolated non-coalescent segments'';
they have also been called ``trapped unary spans'' \citep[by][]{wong2024general}.
Such isolated segments will not be recovered by our algorithm,
and would likely be unrecoverable by any other method.
So, after simulation, we first remove these isolated, non-coalescent segments.
To give an idea of what proportion of the full spans of ancestral nodes
these isolated non-coalescent segments represent,
a simulation of 1000 samples
with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
has about half the total span of all nodes in isolated, non-coalescent segments.
For more discussion of these segments, see \citet{baumdicker2021efficient}.
